Detailed structure of negatively charged membranes of phosphatidylserine and phosphatidylcholine at concentrations of calcium, sodium and potassium salts from molecular dynamics simulations with electronic polarization

### 2019-05-23 Manuscript/manuscript.tex
update text

Here, we extend the ECC approach also to the negatively charged 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-L-serine (POPS) lipid.
We also complement the available experimental data for the the force field quality
evaluation by measuring the acyl chain C--H bond order parameters of POPS in bilayer using natural abundance $^{13}$C NMR.
In addition to the acyl chain order parameters measured here, the quality of the newly developed ECC-POPS force field parameters is
evaluated using previously published C--H bond order parameters of glycerol backbone and headgroup in various ionic
conditions and lipid molar fractions \cite{roux90,NMRlipidsIV}, as well as X-ray scattering form factors\cite{kucerka14}.
The overall improvement of the force field accuracy upon applying ECC
%not only describes the interactions of PS with ions more accurately, 
%but also improves the structure of the PS head group 
%and its interactions with PC in mixed lipid bilayers. 
%The results
paves the way to more realistic MD simulations 
of both neutral and negatively charged lipid bilayers 
for a wide range of applications.

@@

Here, we apply the ECC to the Amber-based lipid17 parameters \cite{lipid17-future} of POPS lipid with a monovalent negative charge.
Because POPS carries a physical total charge as aqueous ions, we apply the theoretically derived scaling factor for
ions $f_q = 0.75$ \cite{leontyev09} to the partial charges of POPS. 
The lipids with a total charge of -0.75 are also neutralized
by the counterion charges of +0.75 in simulations with ECC-ions \cite{Pluharova2014, kohagen16, martinek17}.
Following our previous work for POPC, we use the scaling factor of $f_\sigma = 0.89$ for the Lennard-Jones~$\sigma$ parameters and
use it to scale the parameters of atoms only in the headgroup, glycerol backbone, and carbonyl regions \cite{melcr18, ECC-POPC_nacl_cacl2_files}.
Further optimization of parameters was not done in this work. 
The generated ECC-POPS parameters are available from Ref.~\citenum{??}
\todo{We need here a permanent citation with DOI.
  Easiest would be to add the parameters into a one of the existing Zenodo repositories containing other simulation parameters.
  We could also put a release of Git into Zenodo, cite that and specify the exact directory where parameters are found,
  but this is more complicated for users to find.
}

@@

The C--H bond order parameters were calculated directly from Eq.~\ref{OPequation}.
The time average was first calculated for each lipid, and
the standard error of the mean from these individual values was then used as the
error estimate \cite{botan15,ollila16,NMRlipidsIV}.
Python program ({\tt scripts/calcOrderParameters.py}) that uses the
MDAnalysis library \cite{agrawal11,gowers16} is available in Ref.~\citenum{MATCHgit}. 
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.
The salt concentrations in buffer before solvating the lipids, reported in the used experimental data set \cite{roux90},
were calculated as $[{\rm salt}]=N_{\rm c} \times[{\rm water}]\,/\,N_{\rm w}$,
where $N_{\rm c}$ is the number of cations in simulation, $[{\rm water}]$\,=\,55.5~M and $N_{\rm w}$
is the number of water molecules in the simulation.
As discussed in our previous work \cite{NMRlipidsIV}, the hydration levels of
multilamellae are expected to be sufficiently similar in the used simulations and reference experiments \cite{roux90}.

@@

The Lipid17/Dang simulation gives larger acyl chain order parameters and smaller area per molecule than the
experimental values (Fig.~\ref{simVSexpNOions_POPS} and table~\ref{tab:apls}),
indicating that this simulation predicts a too compact POPS lipid bilayer. This is also in line with
the discrepancy observed in the X-ray form factor between the simulation and experiment (Fig.~\ref{simVSexpNOions_POPS}).
In previous study \cite{NMRlipidsIV}, the Lipid17 force field with {\AA}qvist ions~\cite{aqvist90}
gave a larger area per molecule (57~\AA$^2$) for POPS being closer to the experimental value,
which is probably due to the weaker counterion binding affinity of {\AA}qvist ions to phospholipids. \cite{catte16, NMRlipidsIV}
However,  {\AA}qvist ions produce known artifacts, 
e.g. they exhibit artificial aggregation of ions at larger salt concentrations \cite{kohagen16, chen07, NMRlipidsIV}.
Therefore, the Dang ions with more realisitic bulk properties are used here to analyze the effect of ECC to the properties of a POPS lipid bilayer. 
The ECC-POPS simulation gives a good agreement with the X-ray scattering form factors and experimental
acyl chain order parameters (see Fig.~\ref{simVSexpNOions_POPS}) indicating that the bilayer dimensions and acyl chain
conformations are well described by the force field. The area per lipid from ECC-POPS simulation is slightly
smaller than the value reported from SDP model \cite{kucerka14} (table~\ref{tab:apls}), but the values
agree within their error estimates.

>>>>

Here, we extend the ECC approach also to the negatively charged 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-L-serine (POPS) lipid.
We also complement the available experimental data for the force field quality
evaluation by measuring the acyl chain C--H bond order parameters of POPS in bilayer using natural abundance $^{13}$C NMR.
In addition to the acyl chain order parameters measured here, the quality of the newly developed ECC-POPS force field parameters is
evaluated using previously published C--H bond order parameters of glycerol backbone and headgroup in various ionic
conditions and lipid molar fractions \cite{roux90,NMRlipidsIV}, as well as X-ray scattering form factors\cite{kucerka14}.
The overall improvement of the force field accuracy upon applying ECC
%not only describes the interactions of PS with ions more accurately, 
%but also improves the structure of the PS head group 
%and its interactions with PC in mixed lipid bilayers. 
%The results
paves the way to more realistic MD simulations 
of both neutral and negatively charged lipid bilayers 
for a wide range of applications.

@@

Here, we apply the ECC to the Amber-based lipid17 parameters \cite{lipid17-future} of POPS lipid with a monovalent negative charge.
Because POPS carries a physical total charge as aqueous ions, we apply the theoretically derived scaling factor for
ions $f_q = 0.75$ \cite{leontyev09} to the partial charges of POPS. 
The lipids with a total charge of -0.75 are also neutralized
by the counterion charges of +0.75 in simulations with ECC-ions \cite{Pluharova2014, kohagen16, martinek17}.
Following our previous work for POPC, we use the scaling factor of $f_\sigma = 0.89$ for the Lennard-Jones~$\sigma$ parameters and
use it to scale the parameters of atoms only in the headgroup, glycerol backbone, and carbonyl regions \cite{melcr18}.
Further optimization of parameters was not done in this work. 
The generated ECC-POPS parameters are available from Ref.~\citenum{??}
\todo{We need here a permanent citation with DOI.
  Easiest would be to add the parameters into a one of the existing Zenodo repositories containing other simulation parameters.
  We could also put a release of Git into Zenodo, cite that and specify the exact directory where parameters are found,
  but this is more complicated for users to find.
}

@@

The C--H bond order parameters were calculated directly from Eq.~\ref{OPequation}.
The time average was first calculated for each lipid, and
the standard error of the mean from these individual values was then used as the
error estimate \cite{botan15,ollila16,NMRlipidsIV}.
Python program that uses the
MDAnalysis library \cite{agrawal11,gowers16} is available in Ref.~\citenum{MATCHgit} ({\tt scripts/calcOrderParameters.py}). 
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.
The salt concentrations in buffer before solvating the lipids, reported in the used experimental data set \cite{roux90},
were calculated as $[{\rm salt}]=N_{\rm c} \times[{\rm water}]\,/\,N_{\rm w}$,
where $N_{\rm c}$ is the number of cations in simulation, $[{\rm water}]$\,=\,55.5~M and $N_{\rm w}$
is the number of water molecules in the simulation.
As discussed in our previous work \cite{NMRlipidsIV}, the hydration levels of
multilamellae are expected to be sufficiently similar in the used simulations and reference experiments \cite{roux90}.

@@

The Lipid17/Dang simulation gives 
discrepancies with the experimental X-ray form factor,
larger acyl chain order parameters,
and smaller area per molecule than the experimental values (Fig.~\ref{simVSexpNOions_POPS} and table~\ref{tab:apls}),
indicating that this simulation predicts a too compact POPS lipid bilayer. 
The ECC-POPS simulation gives better agreement with experiments for the X-ray scattering form factors and 
acyl chain order parameters (Fig.~\ref{simVSexpNOions_POPS}),
indicating that the bilayer dimensions and acyl chain
conformations are well described by the force field. The area per lipid from ECC-POPS simulation is slightly
smaller than the value reported from SDP model \cite{kucerka14} (table~\ref{tab:apls}), but the values
agree within their error estimates. The larger area in ECC-POPS simulation can be
explained by increased headgroup repulsion due to lower counterion binding affinity (Fig. \ref{??}).
This explains also the larger area per molecule (57~\AA$^2$) reported for Lipid17 POPS simulations
%being closer to the experimental value, 
%In previous study \cite{NMRlipidsIV}, the Lipid17 force field
with {\AA}qvist ions~\cite{NMRlipidsIV}.
%gave a 
%which is probably due to the weaker counterion binding affinity of {\AA}qvist ions to phospholipids. \cite{catte16, NMRlipidsIV}
However, the Dang ions %with more realisitic bulk properties
are used here to analyze the effect of ECC to the properties of a POPS lipid bilayer because
{\AA}qvist ions produce known artifacts, like artificial aggregation of ions at larger salt concentrations \cite{kohagen16, chen07, NMRlipidsIV}.
